---
title: "Machine Learning in R"
author: "K Mingoulin"
date: "12/5/2019"
output:  
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    number_sections: no
    theme: readable
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
#csl: Bib_Citation_Files/the-isme-journal.csl
#bibliography: Bib_Citation_Files/2016-05-09-R-packages.bib
---

<!--BEGIN:  Set the global options and load packages-->

```{r set-global-options, echo = FALSE}
knitr::opts_chunk$set(eval = TRUE, 
                      echo = TRUE, 
                      cache = FALSE,
                      include = TRUE,
                      collapse = FALSE,
                      dependson = NULL,
                      engine = "R", # Chunks will always have R code, unless noted
                      error = TRUE,
                      fig.path="Figures/",  # Set the figure options
                      fig.align = "center", 
                      fig.width = 7,
                      fig.height = 7)
```

******************************************************************************************

# Introduction

Welcome to my blog. If you are reading this, most likely you've heard or have taken an amizing course taught by **Andrew Ng** on *Machine Learning*. I had the privelege of taking this course in 2011 and was able to rely on the learnings from it for years to come. One thing I always wished for is that course was taught in **R** instead of *Octave/Matlab*, so I decided to do all the assignments in the course using **R**.  

I will make every effort to avoid using any packages other that **ggplot2**. Enjoy!

The **Github repository** for this blog can be found [here](https://github.com/kostya75/ML2020). 

# Week 2: Gradient Descent

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
